---
title: "Old Kappa-Conditioned vs Revised Kappa-Collapsed Sampler"
subtitle: "Collaborator-Facing Diagnostic Comparison"
author: "MSSTNB simulation diagnostics"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
format:
  pdf:
    toc: true
    number-sections: true
    geometry: margin=1in
    include-in-header:
      text: |
        \usepackage{booktabs}
        \usepackage{float}
        \usepackage{placeins}
        \usepackage{array}
        \hyphenpenalty=10000
        \exhyphenpenalty=10000
        \sloppy
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
root <- "."

path <- function(...) file.path(root, ...)

compare_dir <- path("analysis_collab_diagnostics", "old_vs_revised_compare")

required_files <- c(
  compact = path(compare_dir, "COMPARE4_old_vs_revised_common_success_compact.csv"),
  flags = path(compare_dir, "COMPARE4_interpretation_flags.csv"),
  old_common = path(compare_dir, "COMPARE4_old_metrics_common_success_by_scenario.csv"),
  revised_common = path(compare_dir, "COMPARE4_revised_metrics_common_success_by_scenario.csv"),
  old_failure = path("analysis_collab_diagnostics", "old_sampler_full_metrics", "OLD7_old_full_failure_warning_summary.csv"),
  revised_failure = path(compare_dir, "COMPARE3_revised_failure_status_summary.csv")
)

missing_files <- required_files[!file.exists(required_files)]
if (length(missing_files) > 0) {
  stop(
    "The following required files are missing. Run OLD-7 and COMPARE-3/COMPARE-4 first:\n",
    paste(missing_files, collapse = "\n")
  )
}

compact <- read.csv(required_files[["compact"]], stringsAsFactors = FALSE)
flags <- read.csv(required_files[["flags"]], stringsAsFactors = FALSE)
old_common <- read.csv(required_files[["old_common"]], stringsAsFactors = FALSE)
revised_common <- read.csv(required_files[["revised_common"]], stringsAsFactors = FALSE)
old_failure <- read.csv(required_files[["old_failure"]], stringsAsFactors = FALSE)
revised_failure <- read.csv(required_files[["revised_failure"]], stringsAsFactors = FALSE)

clean_nan_df <- function(x) {
  for (nm in names(x)) {
    if (is.numeric(x[[nm]])) {
      x[[nm]][is.nan(x[[nm]])] <- NA_real_
    }
  }
  x
}

compact <- clean_nan_df(compact)
flags <- clean_nan_df(flags)
old_common <- clean_nan_df(old_common)
revised_common <- clean_nan_df(revised_common)

fmt <- function(x, digits = 4) {
  # Avoid displaying tiny numerical noise as -0.0000.
  x_num <- suppressWarnings(as.numeric(x))
  threshold <- 0.5 * 10^(-digits)
  x_num[is.finite(x_num) & abs(x_num) < threshold] <- 0

  ifelse(
    is.na(x_num),
    "NA",
    formatC(x_num, digits = digits, format = "f")
  )
}

fmt_pct <- function(x, digits = 1) {
  ifelse(
    is.na(x),
    "NA",
    paste0(formatC(100 * x, digits = digits, format = "f"), "%")
  )
}

as_display_table <- function(df, digits = 4, percent_cols = character()) {
  out <- df
  for (nm in names(out)) {
    if (nm %in% percent_cols) {
      out[[nm]] <- fmt_pct(out[[nm]])
    } else if (is.numeric(out[[nm]])) {
      out[[nm]] <- fmt(out[[nm]], digits = digits)
    }
  }
  out
}

kable_clean <- function(df, caption = NULL, align = NULL, escape = TRUE) {
  if (is.null(align)) {
    align <- paste0("l", paste(rep("c", ncol(df) - 1), collapse = ""))
  }
  knitr::kable(
    df,
    caption = caption,
    align = align,
    booktabs = TRUE,
    escape = escape,
    row.names = FALSE
  )
}

make_rmse_long <- function(df, prefix, label) {
  do.call(
    rbind,
    lapply(seq_len(nrow(df)), function(i) {
      data.frame(
        Scenario = df$scenario[i],
        Quantity = label,
        Old = df[[paste0(prefix, "_old")]][i],
        Revised = df[[paste0(prefix, "_revised_common_reps")]][i],
        Delta = df[[paste0("delta_", prefix, "_revised_minus_old")]][i],
        stringsAsFactors = FALSE
      )
    })
  )
}

knitr::opts_chunk$set(
  fig.align = "center",
  fig.width = 6.5,
  fig.height = 3.2,
  out.width = "95%"
)

options(knitr.kable.NA = "NA")

s4a <- compact[compact$scenario == "S4A", ]
non_s4a <- compact[compact$scenario != "S4A", ]

max_abs_baseline_delta_non_s4a <- max(
  abs(non_s4a$delta_baseline_logrisk_rmse_revised_minus_old),
  na.rm = TRUE
)

max_abs_ppc95_delta <- max(
  abs(compact$delta_ppc95_revised_minus_old),
  na.rm = TRUE
)

max_abs_assoc_delta_non_s1 <- max(
  abs(compact$delta_abs_logkappa_eta_cor_revised_minus_old[compact$scenario != "S1"]),
  na.rm = TRUE
)
```

# Executive summary

The revised kappa-collapsed sampler answers the collaborator's concern favorably. It preserves posterior predictive calibration and produces nearly identical regression and baseline-risk recovery to the original sampler in S1--S3 and S4B--S4E. Its main advantage is robustness in S4A, the sparse-count stress test: the original sampler has one baseline-analysis failure out of ten replicates, whereas the revised sampler has no baseline-analysis failures.

On the common-success S4A replicates, the revised sampler reduces baseline log-risk RMSE from `r fmt(s4a$baseline_logrisk_rmse_old)` to `r fmt(s4a$baseline_logrisk_rmse_revised_common_reps)` and log-lambda RMSE from `r fmt(s4a$loglambda_rmse_old)` to `r fmt(s4a$loglambda_rmse_revised_common_reps)`, while PPC 95% coverage is essentially unchanged (`r fmt(s4a$ppc_coverage95_old)` versus `r fmt(s4a$ppc_coverage95_revised_common_reps)`).

# Purpose

This report compares two implementations of the MSSTNB sampler:

1. **Original kappa-conditioned sampler.** The regression and spatial baseline update is performed conditional on sampled $\kappa$.
2. **Revised kappa-collapsed sampler.** The baseline regression and spatial update is based on the marginal negative-binomial likelihood, with $\kappa$ sampled later for the dispersion and dynamic-risk components.

The collaborator's question is whether the revised sampler removes baseline distortion or numerical fragility without degrading predictive performance. We therefore report four families of diagnostics:

- regression recovery for $\beta_0$, $\beta_1$, and $\beta_2$;
- recovery of the fitted baseline log-risk $\eta_{tj} = \beta_0 + \beta_1 x_{1,tj} + \beta_2 x_{2,tj} + \phi_j$;
- recovery of the dynamic multiplier on the log scale, $\log \lambda_{tj}$;
- posterior predictive calibration and the diagnostic association between $\log(\kappa_{tj})$ and the fitted baseline log-risk.

All comparisons use the same formal MCMC settings: 20,000 iterations, 10,000 burn-in iterations, and thinning by 5. The formal runs are single-chain, so formal R-hat is not available without a separate multi-chain rerun.

# Why common-success comparison is needed

The original sampler had one baseline-analysis failure in scenario S4A. Comparing all revised fits against only successful old fits would mix different replicate sets in S4A. Therefore, the main metric comparison below is a **common-success comparison**: for each scenario, the revised sampler is restricted to the same replicate set for which the old sampler was baseline-usable. Failure rates are reported separately.

```{r status-table}
status_table <- compact[, c(
  "scenario",
  "n_total",
  "n_common_success",
  "old_n_analysis_failed",
  "old_failure_rate",
  "revised_n_non_ok"
)]

names(status_table) <- c(
  "Scen.",
  "Total",
  "Common success",
  "Old failures",
  "Old failure rate",
  "Revised non-OK"
)

status_table <- as_display_table(status_table, digits = 0, percent_cols = c("Old failure rate", "Old fail rate"))

kable_clean(
  status_table,
  caption = "Baseline-analysis status used to define the common-success replicate set."
)
```

The revised sampler completed all 110 fits used in this diagnostic set without baseline-analysis failures. The original sampler was baseline-usable in all scenarios except S4A, where it had `r s4a$old_n_analysis_failed` failure out of `r s4a$n_total` replicates, corresponding to a failure rate of `r fmt_pct(s4a$old_failure_rate)`.

# Regression recovery

```{r regression-rmse-table}
reg_table <- do.call(
  rbind,
  list(
    data.frame(
      Scenario = compact$scenario,
      Parameter = "$\\beta_0$",
      Old = compact$beta0_rmse_old,
      Revised = compact$beta0_rmse_revised_common_reps,
      Delta = compact$delta_beta0_rmse_revised_minus_old,
      stringsAsFactors = FALSE
    ),
    data.frame(
      Scenario = compact$scenario,
      Parameter = "$\\beta_1$",
      Old = compact$beta1_rmse_old,
      Revised = compact$beta1_rmse_revised_common_reps,
      Delta = compact$delta_beta1_rmse_revised_minus_old,
      stringsAsFactors = FALSE
    ),
    data.frame(
      Scenario = compact$scenario,
      Parameter = "$\\beta_2$",
      Old = compact$beta2_rmse_old,
      Revised = compact$beta2_rmse_revised_common_reps,
      Delta = compact$delta_beta2_rmse_revised_minus_old,
      stringsAsFactors = FALSE
    )
  )
)

reg_table <- reg_table[order(reg_table$Scenario, reg_table$Parameter), ]
reg_table <- as_display_table(reg_table, digits = 4)

kable_clean(
  reg_table,
  caption = "Regression RMSE on common-success replicates. Delta is revised minus old.",
  escape = FALSE
)
```

Across S1--S3 and S4B--S4E, the regression RMSE values are nearly identical between the two samplers. S4A is the only scenario in which stability differs: on the common-success replicates, the revised sampler has slightly smaller $\beta_0$ RMSE than the old sampler (`r fmt(s4a$beta0_rmse_revised_common_reps)` versus `r fmt(s4a$beta0_rmse_old)`).

# Baseline-risk and dynamic-risk recovery

```{r baseline-lambda-table}
baseline_lambda_table <- compact[, c(
  "scenario",
  "baseline_logrisk_rmse_old",
  "baseline_logrisk_rmse_revised_common_reps",
  "delta_baseline_logrisk_rmse_revised_minus_old",
  "loglambda_rmse_old",
  "loglambda_rmse_revised_common_reps",
  "delta_loglambda_rmse_revised_minus_old"
)]

names(baseline_lambda_table) <- c(
  "Scen.",
  "Base RMSE old",
  "Base RMSE rev.",
  "Delta base",
  "log-lambda old",
  "log-lambda rev.",
  "Delta log-lambda"
)

baseline_lambda_table <- as_display_table(baseline_lambda_table, digits = 4)

kable_clean(
  baseline_lambda_table,
  caption = "Baseline log-risk and dynamic log-lambda recovery on common-success replicates. Delta is revised minus old."
)
```

The baseline log-risk recovery is essentially unchanged outside S4A. Across S1--S3 and S4B--S4E, the largest absolute change in baseline log-risk RMSE is only `r fmt(max_abs_baseline_delta_non_s4a)`. In S4A, the revised sampler improves the common-success baseline log-risk RMSE from `r fmt(s4a$baseline_logrisk_rmse_old)` to `r fmt(s4a$baseline_logrisk_rmse_revised_common_reps)`. It also improves the log-lambda RMSE from `r fmt(s4a$loglambda_rmse_old)` to `r fmt(s4a$loglambda_rmse_revised_common_reps)`.

```{r fig-baseline-rmse, fig.cap="Baseline log-risk RMSE by scenario on common-success replicates."}
mat <- rbind(
  Old = compact$baseline_logrisk_rmse_old,
  Revised = compact$baseline_logrisk_rmse_revised_common_reps
)
barplot(
  mat,
  beside = TRUE,
  names.arg = compact$scenario,
  ylab = "Baseline log-risk RMSE",
  legend.text = rownames(mat),
  args.legend = list(x = "topleft", bty = "n")
)
```

```{r fig-loglambda-rmse, fig.cap="Log-lambda RMSE by scenario on common-success replicates."}
mat <- rbind(
  Old = compact$loglambda_rmse_old,
  Revised = compact$loglambda_rmse_revised_common_reps
)
barplot(
  mat,
  beside = TRUE,
  names.arg = compact$scenario,
  ylab = "log-lambda RMSE",
  legend.text = rownames(mat),
  args.legend = list(x = "topleft", bty = "n")
)
```

# Posterior predictive calibration

```{r ppc-table}
ppc_table <- compact[, c(
  "scenario",
  "ppc_coverage95_old",
  "ppc_coverage95_revised_common_reps",
  "delta_ppc95_revised_minus_old"
)]

names(ppc_table) <- c(
  "Scenario",
  "PPC 95% old",
  "PPC 95% revised",
  "Delta PPC 95%"
)

ppc_table <- as_display_table(ppc_table, digits = 4)

kable_clean(
  ppc_table,
  caption = "Posterior predictive cell-level 95% coverage on common-success replicates. Delta is revised minus old."
)
```

Posterior predictive calibration is not degraded by the revised sampler. The largest absolute change in 95% posterior predictive cell coverage is only `r fmt(max_abs_ppc95_delta)`. In S4A, predictive coverage is essentially unchanged: old = `r fmt(s4a$ppc_coverage95_old)`, revised = `r fmt(s4a$ppc_coverage95_revised_common_reps)`.

```{r fig-ppc95, fig.cap="Posterior predictive 95% coverage by scenario on common-success replicates."}
mat <- rbind(
  Old = compact$ppc_coverage95_old,
  Revised = compact$ppc_coverage95_revised_common_reps
)
barplot(
  mat,
  beside = TRUE,
  names.arg = compact$scenario,
  ylab = "PPC 95% coverage",
  ylim = c(0, 1.05),
  legend.text = rownames(mat),
  args.legend = list(x = "bottomright", bty = "n")
)
abline(h = 0.95, lty = 2)
```

# Diagnostic association between $\log(\kappa)$ and baseline log-risk

```{r kappa-association-table}
assoc_table <- compact[, c(
  "scenario",
  "logkappa_eta_abs_cor_mean_old",
  "logkappa_eta_abs_cor_mean_revised_common_reps",
  "delta_abs_logkappa_eta_cor_revised_minus_old"
)]

names(assoc_table) <- c(
  "Scenario",
  "Abs cor old",
  "Abs cor revised",
  "Delta abs cor"
)

assoc_table <- as_display_table(assoc_table, digits = 4)

kable_clean(
  assoc_table,
  caption = "Mean absolute draw-wise correlation between log-kappa and baseline log-risk. Delta is revised minus old."
)
```

The diagnostic association between $\log(\kappa)$ and the fitted baseline log-risk remains weak. Excluding S1, where the revised baseline-only fit does not store posterior $\kappa$ samples and the association is not applicable, the largest absolute change in the mean absolute association is `r fmt(max_abs_assoc_delta_non_s1)`. Thus, the revised sampler does not introduce stronger coupling between $\kappa$ and the baseline log-risk.

# Interpretation flags

```{r flags-table}
flags_display <- data.frame(
  Scen. = flags$scenario,
  `Old fail rate` = flags$old_failure_rate,
  `Base RMSE delta` = flags$baseline_rmse_delta,
  `log-lambda delta` = flags$loglambda_rmse_delta,
  `PPC95 delta` = flags$ppc95_delta,
  `Abs cor delta` = flags$abs_logkappa_eta_cor_delta,
  `Base material` = flags$baseline_rmse_material_change,
  `PPC material drop` = flags$ppc95_material_drop,
  `Old failure` = flags$old_has_failure,
  check.names = FALSE
)

flags_display <- as_display_table(flags_display, digits = 4, percent_cols = c("Old failure rate", "Old fail rate"))

kable_clean(
  flags_display,
  caption = "Automatic interpretation flags. Material baseline RMSE change is defined here as absolute delta greater than 0.01; material PPC drop is defined as PPC 95% coverage delta less than -0.01."
)
```

Only S4A is flagged for a material baseline RMSE change and for old-sampler failure. This is expected because S4A is the sparse-count stress test. Importantly, S4A is not flagged for predictive degradation: the revised sampler preserves posterior predictive calibration while reducing baseline and log-lambda RMSE on the common-success replicate set.

# Answer to the collaborator's question

The revised kappa-collapsed sampler does **not** degrade predictive performance. On common-success replicates, posterior predictive 95% coverage is virtually unchanged across all scenarios. The largest absolute change in PPC 95% coverage is `r fmt(max_abs_ppc95_delta)`, and no scenario shows a material decrease.

The revised sampler also does **not** introduce a stronger diagnostic association between $\log(\kappa)$ and the fitted baseline log-risk. Outside S1, where the revised baseline-only fit does not store $\kappa$ samples, the association remains weak and changes only negligibly.

The main practical advantage of the revised sampler is robustness under sparse counts. The original kappa-conditioned sampler has one baseline-analysis failure in S4A, whereas the revised sampler has no baseline-analysis failures in any scenario. On the common-success S4A replicates, the revised sampler reduces baseline log-risk RMSE from `r fmt(s4a$baseline_logrisk_rmse_old)` to `r fmt(s4a$baseline_logrisk_rmse_revised_common_reps)` and log-lambda RMSE from `r fmt(s4a$loglambda_rmse_old)` to `r fmt(s4a$loglambda_rmse_revised_common_reps)`, while keeping PPC 95% coverage essentially unchanged.

Overall, the revised sampler provides a more stable and target-aligned implementation. Among successful fits, it agrees closely with the original sampler in the non-stress and non-sparse scenarios; under sparse counts, it avoids an old-sampler failure and improves latent baseline/dynamic-risk recovery without sacrificing posterior predictive calibration.

# Files used

This report uses files generated by the diagnostic workflow. The source files are stored under `analysis_collab_diagnostics/old_vs_revised_compare/` and `analysis_collab_diagnostics/old_sampler_full_metrics/`.

```{r files-table}
files_table <- data.frame(
  Role = c(
    "Common-success compact table",
    "Interpretation flags",
    "Old common-success metrics",
    "Revised common-success metrics",
    "Old failure/warning summary",
    "Revised failure/status summary"
  ),
  File = c(
    "COMPARE4 compact table",
    "COMPARE4 interpretation flags",
    "COMPARE4 old scenario metrics",
    "COMPARE4 revised scenario metrics",
    "OLD7 old failure summary",
    "COMPARE3 revised status summary"
  ),
  stringsAsFactors = FALSE
)

kable_clean(files_table, caption = "Input files used by this report.")
```
