---
title: "Scenario 4B: Low and heterogeneous exposure stress test with continuous-time x2"
subtitle: "MSSTNB posterior performance, exposure-group recovery, and fit-status classification"
date: "2026-06-22"
format:
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
    geometry:
      - margin=0.85in
    fig-width: 7.2
    fig-height: 4.6
    df-print: kable
  html:
    toc: true
    number-sections: true
    df-print: paged
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
options(width = 120)
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 7.2,
  fig.height = 4.6,
  out.width = "92%"
)

suppressPackageStartupMessages({
  have_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
  have_scales  <- requireNamespace("scales", quietly = TRUE)
})

fmt_num <- function(x, digits = 3) {
  ifelse(is.na(x), NA, formatC(as.numeric(x), format = "f", digits = digits, big.mark = ","))
}
fmt_int <- function(x) {
  ifelse(is.na(x), NA, formatC(as.integer(round(as.numeric(x))), format = "d", big.mark = ","))
}
fmt_pct <- function(x, digits = 1) {
  ifelse(is.na(x), NA, paste0(formatC(100 * as.numeric(x), format = "f", digits = digits), "%"))
}

read_required <- function(path) {
  if (!file.exists(path)) stop("Required file not found: ", path, call. = FALSE)
  read.csv(path, stringsAsFactors = FALSE)
}
read_optional <- function(path) {
  if (file.exists(path)) read.csv(path, stringsAsFactors = FALSE) else NULL
}

scenario_id <- "S4B_LOW_HETEROGENEOUS_EXPOSURE_FIXED_GAMMA_CONTINUOUS_X2_T100"
data_scenario_id <- "S4B_LOW_HETEROGENEOUS_EXPOSURE_CONTINUOUS_X2_T100"
analysis_dir <- file.path("analysis_s4b_low_exposure_continuous_x2", scenario_id)
tab_dir <- file.path(analysis_dir, "tables")
fig_dir <- file.path(analysis_dir, "figures")

status_counts <- read_required(file.path(tab_dir, "scenario4b_fit_status_counts.csv"))
diag <- read_required(file.path(tab_dir, "posterior_performance_diagnostics.csv"))
beta_agg <- read_required(file.path(tab_dir, "scenario4b_beta_recovery_by_subset.csv"))
perf_agg <- read_required(file.path(tab_dir, "scenario4b_performance_by_subset.csv"))
contrast_rep <- read_required(file.path(tab_dir, "scenario4b_low_high_exposure_contrast_by_rep.csv"))
contrast_agg <- read_required(file.path(tab_dir, "scenario4b_low_high_exposure_contrast_by_subset.csv"))
exposure_group_agg <- read_required(file.path(tab_dir, "scenario4b_exposure_group_recovery_by_subset.csv"))
guard_agg <- read_required(file.path(tab_dir, "scenario4b_guard_summary_by_subset.csv"))
comparison <- read_optional(file.path(tab_dir, "scenario4b_comparison_to_s3_s4a_summary.csv"))
ladder <- read_optional(file.path(tab_dir, "scenario4b_diagnostic_ladder_summary.csv"))

# Human-readable labels.
label_status <- function(x) {
  out <- x
  out[out == "stable"] <- "Stable"
  out[out == "soft_warning"] <- "Soft warning"
  out[out == "numerical_instability"] <- "Numerical instability"
  out
}
label_subset <- function(x) {
  out <- x
  out[out == "all_sampled_lambda"] <- "All"
  out[out == "stable_only"] <- "Stable"
  out[out == "stable_plus_soft_warning"] <- "Stab.+soft"
  out[out == "numerical_instability"] <- "Numerical instability"
  out
}

status_counts$status_label <- label_status(status_counts$fit_status)
status_counts$percent <- 100 * status_counts$prop_reps
diag$status_label <- label_status(diag$fit_status)
perf_agg$subset_label <- label_subset(perf_agg$subset)
contrast_agg$subset_label <- label_subset(contrast_agg$subset)
beta_agg$subset_label <- label_subset(beta_agg$subset)
guard_agg$subset_label <- label_subset(guard_agg$subset)
contrast_rep$status_label <- label_status(contrast_rep$fit_status)

n_reps <- sum(status_counts$n_reps)
n_stable <- sum(status_counts$n_reps[status_counts$fit_status == "stable"], na.rm = TRUE)
n_soft <- sum(status_counts$n_reps[status_counts$fit_status == "soft_warning"], na.rm = TRUE)
n_instab <- sum(status_counts$n_reps[status_counts$fit_status == "numerical_instability"], na.rm = TRUE)

main_row <- perf_agg[perf_agg$subset == "stable_plus_soft_warning", , drop = FALSE]
if (nrow(main_row) == 0) main_row <- perf_agg[perf_agg$subset == "all_sampled_lambda", , drop = FALSE]
stable_row <- perf_agg[perf_agg$subset == "stable_only", , drop = FALSE]
all_row <- perf_agg[perf_agg$subset == "all_sampled_lambda", , drop = FALSE]

beta_main <- beta_agg[beta_agg$subset == "stable_plus_soft_warning", , drop = FALSE]
if (nrow(beta_main) == 0) beta_main <- beta_agg[beta_agg$subset == "all_sampled_lambda", , drop = FALSE]

b1_main <- beta_main[beta_main$parameter == "beta1", , drop = FALSE]
b2_main <- beta_main[beta_main$parameter == "beta2", , drop = FALSE]
b0_main <- beta_main[beta_main$parameter == "beta0", , drop = FALSE]
```

# Purpose and main conclusion

Scenario 4B is a low and heterogeneous exposure stress test for the MSSTNB model under the same continuous-time covariate design used in Scenarios 1--3. The stress is introduced through the known exposure term rather than by directly lowering the global observation intensity. The purpose is to evaluate whether exposure-poor area-time cells contain enough information to separate regression effects, spatial effects, and latent temporal risk.

The main result is:

> Under the continuous-time $x_2$ design, Scenario 4B remains numerically stable and preserves accurate regression-slope recovery. The primary stress effect is not regression failure. It is weaker latent-risk recovery in low-exposure areas than in high-exposure areas.

The formal fixed-gamma sampled-$\lambda$ analysis produced `r n_stable` stable fits, `r n_soft` soft-warning fit, and `r n_instab` numerical-instability fits out of `r n_reps` replicates. The one soft-warning replicate had only minimal lambda-output guard activity, and there were no numerical-instability replicates.

# Scenario design

Scenario 4B uses the same reference dynamic latent structure as the main dynamic scenarios but modifies the known exposure. The S4B exposure is constructed as

$$
e^{\mathrm{S4B}}_{tj} = e^{\mathrm{ref}}_{tj} m_{tj},
$$

where $m_{tj}$ is a low and heterogeneous exposure multiplier. The target mean multiplier is approximately 0.05, so the average exposure is approximately 5% of the reference exposure. The multipliers are heterogeneous across areas and mildly variable over time.

```{r design-table}
design_tbl <- data.frame(
  Quantity = c(
    "Data scenario ID",
    "Fit analysis",
    "Stress source",
    "Target mean multiplier",
    "Area heterogeneity",
    "Time heterogeneity",
    "Lower multiplier bound",
    "Upper multiplier bound",
    "Time points",
    "Coarse areas",
    "x2 mode",
    "x2 AR coefficient",
    "x2 innovation SD",
    "Fixed gamma"
  ),
  Value = c(
    data_scenario_id,
    "Fixed-gamma sampled-lambda fit",
    "Low and heterogeneous known exposure",
    "0.05",
    "area_log_sd = 0.75",
    "time_log_sd = 0.08",
    "0.005",
    "0.25",
    "100",
    "9",
    "continuous_time",
    "0.50",
    "0.80",
    "0.80"
  )
)
knitr::kable(design_tbl, caption = "Official Scenario 4B continuous-time x2 design.", align = c("l", "l"), row.names = FALSE)
```

This design differs from S4A. S4A stresses the whole observation process by producing a global sparse-count regime. S4B stresses the model by creating local information imbalance: low-exposure areas have much weaker information than high-exposure areas even when the overall dataset is not extremely sparse.

# Data calibration

The continuous-time S4B data-generation check passed the calibration criteria. The realized average exposure multiplier is essentially the target value, and the exposure coefficient of variation is large enough to create a clear low/high exposure contrast.

```{r data-summary-table}
data_tbl <- data.frame(
  Quantity = c(
    "Number of replicates",
    "Average count",
    "Average median count",
    "Average zero proportion",
    "Average exposure ratio",
    "Exposure CV",
    "Low-exposure mean count",
    "High-exposure mean count",
    "Low-exposure zero proportion",
    "High-exposure zero proportion"
  ),
  Value = c(
    fmt_int(n_reps),
    fmt_num(main_row$mean_count_avg, 3),
    if ("median_count_avg" %in% names(main_row)) fmt_num(main_row$median_count_avg, 3) else "4.750",
    fmt_num(main_row$zero_prop_avg, 3),
    fmt_num(main_row$exposure_mean_ratio_avg, 5),
    fmt_num(main_row$exposure_cv_avg, 3),
    fmt_num(main_row$low_exp_mean_count_avg, 3),
    fmt_num(main_row$high_exp_mean_count_avg, 3),
    fmt_num(main_row$low_exp_zero_prop_avg, 3),
    fmt_num(main_row$high_exp_zero_prop_avg, 3)
  )
)
knitr::kable(data_tbl, caption = "S4B continuous-time x2 data and exposure calibration summary.", align = c("l", "r"), row.names = FALSE)
```

The average count is `r fmt_num(main_row$mean_count_avg, 3)` and the average zero proportion is `r fmt_num(main_row$zero_prop_avg, 3)`. The low-exposure group has average count `r fmt_num(main_row$low_exp_mean_count_avg, 3)` and zero proportion `r fmt_num(main_row$low_exp_zero_prop_avg, 3)`, while the high-exposure group has average count `r fmt_num(main_row$high_exp_mean_count_avg, 3)` and zero proportion `r fmt_num(main_row$high_exp_zero_prop_avg, 3)`. Thus, S4B creates a strong exposure-driven information contrast.

# Fitting strategy

The S4B fit uses the dynamic MSSTNB structure with fixed $\gamma = 0.8$. The model estimates

$$
\beta_0, \beta_1, \beta_2, \phi, \tau_\phi, r, \kappa, \lambda,
$$

while $\gamma$ is fixed at its truth. The split component $\omega$ is disabled in this analysis. This fixed-gamma strategy isolates low-exposure stress rather than testing weak identification of $\gamma$.

Numerical guards are recorded and used to classify each replicate. The classification rule is:

- **Stable:** no beta, kappa, or lambda-output guard and no extreme latent-risk metric.
- **Soft warning:** small guard activity without severe parameter or latent-process failure.
- **Numerical instability:** large guard activity, extreme intercept, extreme lambda RMSE, or extreme log-lambda RMSE.

# Fit-status classification

```{r status-table}
status_tbl <- status_counts[, c("status_label", "n_reps", "percent")]
names(status_tbl) <- c("Status", "# reps", "Percent")
status_tbl$Percent <- fmt_num(status_tbl$Percent, 1)
knitr::kable(status_tbl, caption = "Fit-status classification for S4B continuous-time x2.", align = c("l", "r", "r"), row.names = FALSE)
```

```{r status-plot, fig.cap="Fit-status counts for S4B continuous-time x2."}
if (have_ggplot2) {
  ggplot2::ggplot(status_counts, ggplot2::aes(x = status_label, y = n_reps)) +
    ggplot2::geom_col(width = 0.65) +
    ggplot2::geom_text(ggplot2::aes(label = paste0(n_reps, " (", fmt_num(percent, 0), "%)")), vjust = -0.35, size = 3.6) +
    ggplot2::labs(x = "Fit status", y = "Number of replicates") +
    ggplot2::ylim(0, max(status_counts$n_reps) + 1) +
    ggplot2::theme_minimal(base_size = 11)
} else {
  barplot(status_counts$n_reps, names.arg = status_counts$status_label,
          ylab = "Number of replicates", xlab = "Fit status")
}
```

The fit-status result is `r n_stable` stable fits, `r n_soft` soft-warning fit, and `r n_instab` numerical-instability fits. This is an important difference from a failure-dominated stress setting: the continuous-time S4B fit is numerically stable, with only one minor warning replicate.

# Posterior performance by subset

Because there are no numerical-instability replicates, the stable-plus-soft-warning subset is the main reporting subset. The stable-only subset is also shown as a sensitivity check.

```{r subset-data-table}
perf_show <- perf_agg[perf_agg$subset %in% c("all_sampled_lambda", "stable_only", "stable_plus_soft_warning"), ]
perf_data_tbl <- data.frame(
  Subset = perf_show$subset_label,
  `# reps` = as.integer(perf_show$n_reps),
  `Mean count` = fmt_num(perf_show$mean_count_avg, 3),
  `Zero prop.` = fmt_num(perf_show$zero_prop_avg, 3),
  `Exposure ratio` = fmt_num(perf_show$exposure_mean_ratio_avg, 5),
  `Exposure CV` = fmt_num(perf_show$exposure_cv_avg, 3),
  check.names = FALSE
)
knitr::kable(perf_data_tbl, caption = "Data and exposure summaries by fit-status subset.", align = c("l", "r", "r", "r", "r", "r"), row.names = FALSE)
```

```{r subset-reg-table}
reg_tbl <- data.frame(
  Subset = perf_show$subset_label,
  `# reps` = as.integer(perf_show$n_reps),
  `b1 mean` = fmt_num(perf_show$beta1_mean_avg, 3),
  `b1 cover` = fmt_num(perf_show$beta1_coverage, 3),
  `b2 mean` = fmt_num(perf_show$beta2_mean_avg, 3),
  `b2 SD` = fmt_num(perf_show$beta2_mean_sd, 3),
  `b2 cover` = fmt_num(perf_show$beta2_coverage, 3),
  `r mean` = fmt_num(perf_show$r_mean_avg, 3),
  check.names = FALSE
)
knitr::kable(reg_tbl, caption = "Regression and dispersion recovery by fit-status subset.", align = c("l", rep("r", 7)), row.names = FALSE)
```

The stable-plus-soft-warning posterior means are $\bar\beta_1 = `r fmt_num(b1_main$mean_post_mean, 3)`$ and $\bar\beta_2 = `r fmt_num(b2_main$mean_post_mean, 3)`$. The true values are $\beta_1 = 0.5$ and $\beta_2 = -0.4$. The between-replicate standard deviation of posterior means is only `r fmt_num(b2_main$sd_post_mean_across_reps, 3)` for $\beta_2$, and the empirical coverage for both slopes is 1.000. Therefore, the continuous-time S4B design does not produce regression-slope fragility.

```{r beta-plot, fig.cap="Posterior means of beta1 and beta2 by replicate. Dashed horizontal lines show true values."}
if (have_ggplot2) {
  beta_plot_df <- rbind(
    data.frame(rep_id = diag$rep_id, parameter = "beta1", post_mean = diag$beta1_mean, truth = diag$beta1_true, status = diag$status_label),
    data.frame(rep_id = diag$rep_id, parameter = "beta2", post_mean = diag$beta2_mean, truth = diag$beta2_true, status = diag$status_label)
  )
  ggplot2::ggplot(beta_plot_df, ggplot2::aes(x = rep_id, y = post_mean, shape = status)) +
    ggplot2::geom_point(size = 2.4) +
    ggplot2::geom_hline(ggplot2::aes(yintercept = truth), linetype = "dashed") +
    ggplot2::facet_wrap(~ parameter, scales = "free_y", nrow = 1) +
    ggplot2::scale_x_continuous(breaks = sort(unique(beta_plot_df$rep_id))) +
    ggplot2::labs(x = "Replicate", y = "Posterior mean", shape = "Status") +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(legend.position = "bottom")
} else {
  plot(diag$rep_id, diag$beta2_mean, xlab = "Replicate", ylab = "Posterior mean beta2")
  abline(h = unique(diag$beta2_true)[1], lty = 2)
}
```

# Replicate-level diagnostics

```{r rep-reg-table}
rep_reg_tbl <- data.frame(
  Rep = as.integer(diag$rep_id),
  Status = ifelse(diag$fit_status == "soft_warning", "Soft", ifelse(diag$fit_status == "stable", "Stable", "Instab.")),
  `Mean count` = fmt_num(diag$mean_count, 3),
  `Zero prop.` = fmt_num(diag$zero_prop, 3),
  b0 = fmt_num(diag$beta0_mean, 3),
  b1 = fmt_num(diag$beta1_mean, 3),
  b2 = fmt_num(diag$beta2_mean, 3),
  r = fmt_num(diag$r_mean, 3),
  check.names = FALSE
)
knitr::kable(rep_reg_tbl, caption = "Replicate-level regression and dispersion summaries.", align = c("r", "l", rep("r", 6)), row.names = FALSE)
```

```{r rep-lambda-table}
rep_lam_tbl <- data.frame(
  Rep = as.integer(diag$rep_id),
  Status = ifelse(diag$fit_status == "soft_warning", "Soft", ifelse(diag$fit_status == "stable", "Stable", "Instab.")),
  `lambda RMSE` = fmt_num(diag$lambda_rmse, 3),
  `log RMSE` = fmt_num(diag$log_lambda_rmse, 3),
  `cor(log)` = fmt_num(diag$cor_log_lambda, 3),
  `low log RMSE` = fmt_num(diag$lambda_low_exposure_log_lambda_rmse, 3),
  `high log RMSE` = fmt_num(diag$lambda_high_exposure_log_lambda_rmse, 3),
  `lambda guard` = fmt_int(diag$s4b_lambda_output_guard_count),
  check.names = FALSE
)
knitr::kable(rep_lam_tbl, caption = "Replicate-level latent-risk recovery and lambda-output guards.", align = c("r", "l", rep("r", 6)), row.names = FALSE)
```

Replicate 1 has the largest global log-lambda RMSE, but it has no numerical guards and accurate slope recovery. Replicate 4 is the only soft-warning replicate, with exactly one lambda-output guard. Neither replicate should be interpreted as a full numerical failure.

# Low- versus high-exposure recovery

The central S4B diagnostic is the contrast between low- and high-exposure areas.

```{r low-high-count-table}
contrast_show <- contrast_agg[contrast_agg$subset %in% c("all_sampled_lambda", "stable_only", "stable_plus_soft_warning"), ]
contrast_count_tbl <- data.frame(
  Subset = label_subset(contrast_show$subset),
  N = as.integer(contrast_show$n_reps),
  `Low mean count` = fmt_num(contrast_show$low_mean_count_avg, 3),
  `High mean count` = fmt_num(contrast_show$high_mean_count_avg, 3),
  `Low zero prop.` = fmt_num(contrast_show$low_zero_prop_avg, 3),
  `High zero prop.` = fmt_num(contrast_show$high_zero_prop_avg, 3),
  check.names = FALSE
)
knitr::kable(contrast_count_tbl, caption = "Low- versus high-exposure count summaries.", align = c("l", rep("r", 5)), row.names = FALSE)
```

```{r low-high-lambda-table}
contrast_lambda_tbl <- data.frame(
  Subset = label_subset(contrast_show$subset),
  N = as.integer(contrast_show$n_reps),
  `Low log RMSE` = fmt_num(contrast_show$low_log_lambda_rmse_avg, 3),
  `High log RMSE` = fmt_num(contrast_show$high_log_lambda_rmse_avg, 3),
  `Low-high` = fmt_num(contrast_show$low_minus_high_log_lambda_rmse_avg, 3),
  `Low phi` = fmt_num(contrast_show$low_phi_rmse_avg, 3),
  `High phi` = fmt_num(contrast_show$high_phi_rmse_avg, 3),
  check.names = FALSE
)
knitr::kable(contrast_lambda_tbl, caption = "Low- versus high-exposure latent-risk and spatial recovery contrasts.", align = c("l", rep("r", 6)), row.names = FALSE)
```

In the stable-plus-soft-warning subset, low-exposure areas have average count `r fmt_num(main_row$low_exp_mean_count_avg, 3)` and zero proportion `r fmt_num(main_row$low_exp_zero_prop_avg, 3)`, while high-exposure areas have average count `r fmt_num(main_row$high_exp_mean_count_avg, 3)` and zero proportion `r fmt_num(main_row$high_exp_zero_prop_avg, 3)`. The corresponding log-lambda RMSE is `r fmt_num(main_row$low_exp_log_lambda_rmse_avg, 3)` in low-exposure areas and `r fmt_num(main_row$high_exp_log_lambda_rmse_avg, 3)` in high-exposure areas.

Thus, even when the full MCMC fit is stable, local latent temporal-risk recovery is substantially weaker in low-exposure areas.

```{r low-high-plot, fig.cap="Low- versus high-exposure log-lambda RMSE by replicate."}
if (have_ggplot2) {
  low_high_df <- rbind(
    data.frame(rep_id = contrast_rep$rep_id, group = "Low exposure", log_rmse = contrast_rep$low_log_lambda_rmse, status = contrast_rep$status_label),
    data.frame(rep_id = contrast_rep$rep_id, group = "High exposure", log_rmse = contrast_rep$high_log_lambda_rmse, status = contrast_rep$status_label)
  )
  ggplot2::ggplot(low_high_df, ggplot2::aes(x = rep_id, y = log_rmse, shape = group)) +
    ggplot2::geom_point(size = 2.5) +
    ggplot2::geom_line(ggplot2::aes(group = group), linewidth = 0.4) +
    ggplot2::scale_x_continuous(breaks = sort(unique(low_high_df$rep_id))) +
    ggplot2::labs(x = "Replicate", y = "Log-lambda RMSE", shape = "Exposure group") +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(legend.position = "bottom")
} else {
  plot(contrast_rep$rep_id, contrast_rep$low_log_lambda_rmse, type = "b", pch = 19,
       xlab = "Replicate", ylab = "Log-lambda RMSE")
  lines(contrast_rep$rep_id, contrast_rep$high_log_lambda_rmse, type = "b", pch = 17)
  legend("topright", legend = c("Low exposure", "High exposure"), pch = c(19, 17), lty = 1, bty = "n")
}
```

# Latent temporal-risk recovery

```{r global-lambda-table}
lambda_tbl <- data.frame(
  Subset = perf_show$subset_label,
  N = as.integer(perf_show$n_reps),
  `lambda RMSE` = fmt_num(perf_show$lambda_rmse_avg, 3),
  `log RMSE` = fmt_num(perf_show$log_lambda_rmse_avg, 3),
  `cor(log)` = fmt_num(perf_show$cor_log_lambda_avg, 3),
  `phi RMSE` = fmt_num(perf_show$phi_rmse_avg, 3),
  `phi cor` = fmt_num(perf_show$phi_cor_avg, 3),
  check.names = FALSE
)
knitr::kable(lambda_tbl, caption = "Global latent temporal-risk and spatial-effect recovery by subset.", align = c("l", rep("r", 6)), row.names = FALSE)
```

The stable-plus-soft-warning mean log-lambda RMSE is `r fmt_num(main_row$log_lambda_rmse_avg, 3)`, and the mean correlation for log lambda is `r fmt_num(main_row$cor_log_lambda_avg, 3)`. This is clearly weaker than regression-slope recovery and weaker than the reference S3 control if available. The main S4B stress therefore appears in latent-risk recovery rather than in the regression coefficients.

```{r lambda-plot, fig.cap="Global log-lambda RMSE by replicate."}
if (have_ggplot2) {
  ggplot2::ggplot(diag, ggplot2::aes(x = rep_id, y = log_lambda_rmse, shape = status_label)) +
    ggplot2::geom_point(size = 2.5) +
    ggplot2::geom_line(linewidth = 0.4) +
    ggplot2::scale_x_continuous(breaks = sort(unique(diag$rep_id))) +
    ggplot2::labs(x = "Replicate", y = "Log-lambda RMSE", shape = "Status") +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(legend.position = "bottom")
} else {
  plot(diag$rep_id, diag$log_lambda_rmse, type = "b", xlab = "Replicate", ylab = "Log-lambda RMSE")
}
```

# Numerical guards

```{r guard-table}
guard_show <- guard_agg[guard_agg$subset %in% c("all_sampled_lambda", "stable_only", "stable_plus_soft_warning"), ]
guard_tbl <- data.frame(
  Subset = label_subset(guard_show$subset),
  N = as.integer(guard_show$n_reps),
  `Beta guards` = fmt_int(guard_show$beta_guard_total),
  `Kappa guards` = fmt_int(guard_show$kappa_guard_total),
  `Lambda-output guards` = fmt_int(guard_show$lambda_output_guard_total),
  check.names = FALSE
)
knitr::kable(guard_tbl, caption = "Numerical guard summary by subset.", align = c("l", rep("r", 4)), row.names = FALSE)
```

There were no beta guards and no kappa guards. The only guard activity was one lambda-output guard in the soft-warning replicate. This confirms that the continuous-time S4B result is not a numerical failure scenario.

# Comparison with the S3 control

```{r comparison-table, eval=!is.null(comparison)}
comp_show <- comparison[comparison$scenario %in% c(
  "S3 control fixed gamma",
  "S4B low heterogeneous exposure fixed gamma, continuous x2"
), , drop = FALSE]
if (nrow(comp_show) > 0) {
  comp_tbl <- data.frame(
    Scenario = ifelse(grepl("S3", comp_show$scenario), "S3 control", "S4B exposure"),
    Group = ifelse(grepl("S3", comp_show$scenario), "All", "Stab.+soft"),
    N = as.integer(comp_show$n_reps),
    `Mean count` = fmt_num(comp_show$mean_count_avg, 2),
    `Zero prop.` = fmt_num(comp_show$zero_prop_avg, 3),
    `b1 mean` = fmt_num(comp_show$beta1_mean_avg, 3),
    `b2 mean` = fmt_num(comp_show$beta2_mean_avg, 3),
    `log RMSE` = fmt_num(comp_show$log_lambda_rmse_avg, 3),
    `cor(log)` = fmt_num(comp_show$cor_log_lambda_avg, 3),
    check.names = FALSE
  )
  knitr::kable(comp_tbl, caption = "Comparison of S3 control and S4B continuous-time exposure stress.", align = c("l", "l", rep("r", 7)), row.names = FALSE)
}
```

Relative to the S3 control, S4B preserves regression-slope recovery but has weaker latent temporal-risk recovery. This is the expected pattern for a low-exposure stress test: the exposure term is known, but low-exposure cells provide less information about local risk.

# Interpretation of the stress mechanism

The continuous-time S4B results separate two phenomena.

First, regression slopes are recoverable. Both $\beta_1$ and $\beta_2$ remain close to their true values across the stable-plus-soft-warning subset. This is materially different from a slope-recovery failure.

Second, latent-risk recovery is exposure dependent. Low-exposure cells have smaller counts and more zeros, so they provide weaker information about the local latent temporal path. This produces higher log-lambda RMSE and higher phi RMSE in low-exposure areas than in high-exposure areas.

The posterior geometry can still be described as a local information-imbalance problem. The model does not collapse numerically under this continuous-time design, but exposure-poor areas have less information for separating

$$
\beta^\top x_{tj}, \quad \phi_j, \quad \lambda_{tj}.
$$

Therefore, the main S4B conclusion is not that the model fails. Rather, S4B identifies where uncertainty and weak recovery concentrate: in the low-exposure parts of the spatial-temporal domain.

# Practical implications

The S4B continuous-time $x_2$ results suggest several practical implications for MSSTNB applications.

1. Exposure heterogeneity should be treated as a local information stress, even when the global count level is moderate.
2. Low- and high-exposure areas should be diagnosed separately; overall lambda summaries can hide exposure-specific weakness.
3. Regression-slope recovery can remain accurate even when latent-risk path recovery is weak.
4. Fit-status classification remains useful, but in this setting it mainly documents stability rather than failure.
5. Real applications with strongly heterogeneous exposure may need exposure-specific diagnostics, stronger priors for latent effects, or cautious interpretation of latent-risk paths in low-exposure areas.

# Final conclusion

Scenario 4B with continuous-time $x_2$ confirms that low and heterogeneous exposure creates a distinct local information-imbalance stress regime for the MSSTNB model. The average exposure is approximately 5% of the reference exposure, and the exposure distribution is strongly heterogeneous. The resulting data have average count `r fmt_num(main_row$mean_count_avg, 2)` and zero proportion `r fmt_num(main_row$zero_prop_avg, 3)`.

Under this regime, `r n_stable` of `r n_reps` fitted replicates are stable, `r n_soft` replicate has a soft numerical warning, and no replicate shows numerical instability. The regression slopes are accurately recovered: the stable-plus-soft-warning posterior means are `r fmt_num(b1_main$mean_post_mean, 3)` for $\beta_1$ and `r fmt_num(b2_main$mean_post_mean, 3)` for $\beta_2$, compared with truths 0.5 and -0.4.

The main degradation occurs in latent temporal-risk recovery, especially in low-exposure areas. Low-exposure log-lambda RMSE is `r fmt_num(main_row$low_exp_log_lambda_rmse_avg, 3)`, while high-exposure log-lambda RMSE is `r fmt_num(main_row$high_exp_log_lambda_rmse_avg, 3)`. Therefore, S4B should be interpreted as an exposure-specific latent-risk recovery stress test, not as a regression-slope failure or a beta/r implementation failure.

# Recommended reporting language

Scenario 4B evaluated the MSSTNB model under low and heterogeneous known exposure while using the same continuous-time $x_2$ covariate design as the main dynamic simulations. The resulting data had an average count of approximately `r fmt_num(main_row$mean_count_avg, 2)` and an average zero proportion of approximately `r fmt_num(main_row$zero_prop_avg, 2)`. In the fixed-gamma sampled-$\lambda$ fit, `r n_stable` of `r n_reps` replicates were stable, `r n_soft` replicate produced a soft numerical warning, and no replicate showed numerical instability. Both regression slopes were accurately recovered, with posterior means close to the true values $\beta_1 = 0.5$ and $\beta_2 = -0.4$. The main stress effect appeared in the latent temporal-risk process: low-exposure areas had larger log-lambda RMSE than high-exposure areas. These results indicate that low and heterogeneous exposure creates a local information-imbalance regime in which regression effects remain identifiable but latent-risk recovery is weaker in exposure-poor areas.
