---
title: "Scenario 4C: Strong overdispersion stress test with continuous-time x2"
subtitle: "MSSTNB posterior performance, dispersion recovery, kappa recovery, and fit-status classification"
date: "2026-06-22"
format:
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
    geometry:
      - margin=0.85in
    fig-width: 7.2
    fig-height: 4.4
    df-print: kable
  html:
    toc: true
    number-sections: true
    df-print: paged
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
options(width = 120)
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 7.2,
  fig.height = 4.4,
  out.width = "92%"
)

suppressPackageStartupMessages({
  have_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
})

fmt_num <- function(x, digits = 3) {
  ifelse(is.na(x), NA, formatC(as.numeric(x), format = "f", digits = digits, big.mark = ","))
}
fmt_int <- function(x) {
  ifelse(is.na(x), NA, formatC(as.integer(round(as.numeric(x))), format = "d", big.mark = ","))
}
fmt_pct <- function(x, digits = 1) {
  ifelse(is.na(x), NA, paste0(formatC(100 * as.numeric(x), format = "f", digits = digits), "%"))
}

read_required <- function(path) {
  if (!file.exists(path)) stop("Required file not found: ", path, call. = FALSE)
  read.csv(path, stringsAsFactors = FALSE)
}
read_optional <- function(path) {
  if (file.exists(path)) read.csv(path, stringsAsFactors = FALSE) else NULL
}

scenario_id <- "S4C_STRONG_OVERDISPERSION_FIXED_GAMMA_CONTINUOUS_X2_T100"
data_scenario_id <- "S4C_STRONG_OVERDISPERSION_CONTINUOUS_X2_T100"
analysis_dir <- file.path("analysis_s4c_overdispersion_continuous_x2", scenario_id)
tab_dir <- file.path(analysis_dir, "tables")
fig_dir <- file.path(analysis_dir, "figures")

status_counts <- read_required(file.path(tab_dir, "s4c_fit_status_counts_continuous_x2.csv"))
rep_summary <- read_required(file.path(tab_dir, "s4c_replicate_level_summary_continuous_x2.csv"))
beta_agg <- read_required(file.path(tab_dir, "posterior_beta_aggregate_continuous_x2.csv"))
perf_agg <- read_required(file.path(tab_dir, "posterior_performance_by_subset_continuous_x2.csv"))
r_agg <- read_required(file.path(tab_dir, "r_recovery_by_subset_continuous_x2.csv"))
kappa_agg <- read_required(file.path(tab_dir, "kappa_recovery_by_subset_continuous_x2.csv"))
guard_agg <- read_required(file.path(tab_dir, "numerical_guard_summary_by_subset_continuous_x2.csv"))
comparison <- read_optional(file.path(tab_dir, "scenario_comparison_s3_s4c_continuous_x2.csv"))

label_status <- function(x) {
  out <- x
  out[out == "stable"] <- "Stable"
  out[out == "soft_warning"] <- "Soft warning"
  out[out == "numerical_instability"] <- "Numerical instability"
  out
}
label_subset <- function(x) {
  out <- x
  out[out == "all_sampled_lambda"] <- "All"
  out[out == "stable_only"] <- "Stable"
  out[out == "stable_plus_soft_warning"] <- "Stable+soft"
  out[out == "numerical_instability"] <- "Instab."
  out
}

status_counts$status_label <- label_status(status_counts$fit_status)
status_counts$percent <- 100 * status_counts$prop_reps
rep_summary$status_label <- label_status(rep_summary$fit_status)
perf_agg$subset_label <- label_subset(perf_agg$subset)
r_agg$subset_label <- label_subset(r_agg$subset)
kappa_agg$subset_label <- label_subset(kappa_agg$subset)
guard_agg$subset_label <- label_subset(guard_agg$subset)
beta_agg$subset_label <- label_subset(beta_agg$subset)

n_reps <- sum(status_counts$n_reps)
n_stable <- sum(status_counts$n_reps[status_counts$fit_status == "stable"], na.rm = TRUE)
n_soft <- sum(status_counts$n_reps[status_counts$fit_status == "soft_warning"], na.rm = TRUE)
n_instab <- sum(status_counts$n_reps[status_counts$fit_status == "numerical_instability"], na.rm = TRUE)

main_row <- perf_agg[perf_agg$subset == "stable_plus_soft_warning", , drop = FALSE]
if (nrow(main_row) == 0) main_row <- perf_agg[perf_agg$subset == "all_sampled_lambda", , drop = FALSE]

r_main <- r_agg[r_agg$subset == "stable_plus_soft_warning", , drop = FALSE]
if (nrow(r_main) == 0) r_main <- r_agg[r_agg$subset == "all_sampled_lambda", , drop = FALSE]

kappa_main <- kappa_agg[kappa_agg$subset == "stable_plus_soft_warning", , drop = FALSE]
if (nrow(kappa_main) == 0) kappa_main <- kappa_agg[kappa_agg$subset == "all_sampled_lambda", , drop = FALSE]

beta_main <- beta_agg[beta_agg$subset == "stable_plus_soft_warning", , drop = FALSE]
if (nrow(beta_main) == 0) beta_main <- beta_agg[beta_agg$subset == "all_sampled_lambda", , drop = FALSE]
b1_main <- beta_main[beta_main$parameter == "beta1", , drop = FALSE]
b2_main <- beta_main[beta_main$parameter == "beta2", , drop = FALSE]

rep_order <- rep_summary[order(rep_summary$rep_id), , drop = FALSE]
```

# Purpose and main conclusion

Scenario 4C is a strong overdispersion stress test for the MSSTNB model under the same continuous-time covariate design used in Scenarios 1--3. It preserves the reference mean, exposure, spatial, and temporal latent-risk structure, but reduces the negative-binomial dispersion parameter from $r=15$ to $r=3$.

The main result is:

> Under the continuous-time $x_2$ design, Scenario 4C remains numerically stable in all ten replicates. Strong overdispersion increases count variability and kappa-level variability, but the fitted model preserves accurate regression-slope recovery, learns the smaller dispersion parameter, and recovers the latent temporal-risk path reasonably well.

The formal fixed-gamma sampled-$\lambda$ analysis produced `r n_stable` stable fits, `r n_soft` soft-warning fits, and `r n_instab` numerical-instability fits out of `r n_reps` replicates. There were no beta, kappa, or lambda-output numerical guards.

# Scenario design

The negative-binomial model is parameterized so that

$$
Y_{tj} \sim \mathrm{NB}(\mu_{tj}, r),
$$

with approximate variance

$$
\mathrm{Var}(Y_{tj}) = \mu_{tj} + \frac{\mu_{tj}^2}{r}.
$$

Thus, smaller $r$ means stronger overdispersion. Scenario 4C changes the truth from $r=15$ to $r=3$ while retaining the reference dynamic structure and continuous-time $x_2$ covariate design.

```{r design-table}
design_tbl <- data.frame(
  Quantity = c(
    "Data scenario ID",
    "Fit analysis",
    "Stress source",
    "Reference dispersion",
    "Stress dispersion",
    "Ratio to reference",
    "Exposure",
    "Mean structure",
    "Time points",
    "Coarse areas",
    "x2 mode",
    "x2 AR coefficient",
    "x2 innovation SD",
    "Fixed gamma"
  ),
  Value = c(
    data_scenario_id,
    "Fixed-gamma sampled-lambda fit",
    "Strong NB overdispersion",
    "r = 15",
    "r = 3",
    "0.20",
    "Same as reference dynamic scenario",
    "Same beta/phi/lambda structure as reference dynamic scenario",
    "100",
    "9",
    "continuous_time",
    "0.50",
    "0.80",
    "0.80"
  )
)
knitr::kable(design_tbl, caption = "Official Scenario 4C continuous-time x2 design.", row.names = FALSE)
```

This design differs from S4A and S4B. S4A creates a global sparse-count regime. S4B creates local information imbalance through low and heterogeneous exposure. S4C instead keeps counts abundant but makes observation-level variability substantially larger through the smaller $r$.

# Data calibration

The continuous-time S4C data-generation check passed the calibration criteria. The average count remains high and the zero proportion remains very low, so S4C is not a sparse-count scenario. The stress appears in larger count CV, variance-to-mean ratio, and kappa variability relative to the matched reference $r=15$ setting.

```{r calibration-table}
cal_tbl <- data.frame(
  Quantity = c(
    "Number of replicates",
    "Average count",
    "Average zero proportion",
    "Total count",
    "Maximum observed count",
    "Count CV",
    "Reference count CV",
    "Count CV increase",
    "Variance/mean",
    "Reference variance/mean",
    "Variance/mean increase",
    "Truth kappa CV",
    "Reference kappa CV",
    "Kappa CV increase"
  ),
  Value = c(
    fmt_int(n_reps),
    fmt_num(main_row$mean_count_avg),
    fmt_num(main_row$zero_prop_avg),
    fmt_int(main_row$total_count_sum),
    fmt_int(main_row$max_count_max),
    fmt_num(main_row$count_cv_avg),
    fmt_num(main_row$reference_count_cv_avg),
    fmt_num(main_row$count_cv_increase_avg),
    fmt_num(main_row$variance_to_mean_avg),
    fmt_num(main_row$reference_variance_to_mean_avg),
    fmt_num(main_row$variance_to_mean_increase_avg),
    fmt_num(main_row$kappa_truth_cv_avg),
    fmt_num(main_row$reference_kappa_cv_avg),
    fmt_num(main_row$kappa_cv_increase_avg)
  )
)
knitr::kable(cal_tbl, caption = "S4C continuous-time x2 data and overdispersion calibration summary.", row.names = FALSE)
```

The average count is `r fmt_num(main_row$mean_count_avg)` and the average zero proportion is only `r fmt_num(main_row$zero_prop_avg)`. The kappa truth CV is `r fmt_num(main_row$kappa_truth_cv_avg)`, compared with reference kappa CV `r fmt_num(main_row$reference_kappa_cv_avg)`. Thus, S4C successfully creates an overdispersion stress while preserving abundant count information.

# Fitting strategy

The S4C fit uses the dynamic MSSTNB structure with fixed $\gamma = 0.8$. The model estimates

$$
\beta_0, \beta_1, \beta_2, \phi, \tau_\phi, r, \kappa, \lambda,
$$

while $\gamma$ is fixed at its truth. The split component $\omega$ is disabled in this analysis. This fixed-gamma strategy isolates strong-overdispersion stress rather than re-testing weak identification of $\gamma$.

Numerical guards are recorded and used to classify each replicate. The classification rule is:

- Stable: no beta, kappa, or lambda-output guard and no extreme latent-risk metric.
- Soft warning: small guard activity without severe parameter or latent-process failure.
- Numerical instability: large guard activity, extreme coefficient behavior, extreme lambda RMSE, or extreme log-lambda RMSE.

# Fit-status classification

```{r status-table}
status_tbl <- data.frame(
  Status = status_counts$status_label,
  `# reps` = as.integer(status_counts$n_reps),
  Percent = fmt_num(status_counts$percent, 1),
  check.names = FALSE
)
knitr::kable(status_tbl, caption = "Fit-status classification for S4C continuous-time x2.", row.names = FALSE)
```

```{r status-plot, fig.cap="Fit-status counts for S4C continuous-time x2."}
if (have_ggplot2) {
  ggplot2::ggplot(status_counts, ggplot2::aes(x = status_label, y = n_reps)) +
    ggplot2::geom_col(width = 0.65) +
    ggplot2::geom_text(ggplot2::aes(label = paste0(n_reps, " (", round(percent), "%)")), vjust = -0.25, size = 3.5) +
    ggplot2::scale_y_continuous(limits = c(0, max(status_counts$n_reps) + 1), breaks = seq(0, n_reps, by = 2)) +
    ggplot2::labs(x = "Fit status", y = "Number of replicates") +
    ggplot2::theme_bw(base_size = 11)
}
```

All ten formal fits are stable. This is an important difference from the earlier stress scenarios: under the continuous-time $x_2$ design, strong overdispersion does not create a numerical-failure scenario.

# Posterior performance

Because all replicates are stable, the all-sampled, stable-only, and stable-plus-soft-warning subsets are identical. The tables below therefore report the main stable set directly.

```{r regression-table}
reg_tbl <- data.frame(
  Parameter = c("beta0", "beta1", "beta2", "r"),
  Truth = c(
    fmt_num(beta_main$mean_truth[beta_main$parameter == "beta0"]),
    fmt_num(beta_main$mean_truth[beta_main$parameter == "beta1"]),
    fmt_num(beta_main$mean_truth[beta_main$parameter == "beta2"]),
    fmt_num(r_main$r_truth_avg)
  ),
  `Posterior mean` = c(
    fmt_num(beta_main$mean_post_mean[beta_main$parameter == "beta0"]),
    fmt_num(beta_main$mean_post_mean[beta_main$parameter == "beta1"]),
    fmt_num(beta_main$mean_post_mean[beta_main$parameter == "beta2"]),
    fmt_num(r_main$r_mean_avg)
  ),
  `Across-rep SD` = c(
    fmt_num(beta_main$sd_post_mean_across_reps[beta_main$parameter == "beta0"]),
    fmt_num(beta_main$sd_post_mean_across_reps[beta_main$parameter == "beta1"]),
    fmt_num(beta_main$sd_post_mean_across_reps[beta_main$parameter == "beta2"]),
    fmt_num(r_main$r_mean_sd)
  ),
  Coverage = c(
    fmt_num(beta_main$coverage[beta_main$parameter == "beta0"]),
    fmt_num(beta_main$coverage[beta_main$parameter == "beta1"]),
    fmt_num(beta_main$coverage[beta_main$parameter == "beta2"]),
    fmt_num(r_main$r_region_coverage_95_avg)
  ),
  check.names = FALSE
)
knitr::kable(reg_tbl, caption = "Regression and dispersion recovery for S4C continuous-time x2.", row.names = FALSE)
```

The posterior means for the two slopes are $\bar\beta_1 = `r fmt_num(b1_main$mean_post_mean)`$ and $\bar\beta_2 = `r fmt_num(b2_main$mean_post_mean)`$, compared with truths $\beta_1 = 0.5$ and $\beta_2 = -0.4$. The empirical coverage is `r fmt_num(b1_main$coverage)` for $\beta_1$ and `r fmt_num(b2_main$coverage)` for $\beta_2$. Therefore, strong overdispersion under this continuous-time design does not produce regression-slope fragility.

```{r beta-plot, fig.cap="Posterior means of beta1 and beta2 by replicate. Dashed horizontal lines show true values."}
if (have_ggplot2) {
  beta_plot <- rbind(
    data.frame(rep_id = rep_order$rep_id, parameter = "beta1", post_mean = rep_order$beta1_mean, truth = 0.5, status = rep_order$status_label),
    data.frame(rep_id = rep_order$rep_id, parameter = "beta2", post_mean = rep_order$beta2_mean, truth = -0.4, status = rep_order$status_label)
  )
  ggplot2::ggplot(beta_plot, ggplot2::aes(x = rep_id, y = post_mean)) +
    ggplot2::geom_point(ggplot2::aes(shape = status), size = 2.1) +
    ggplot2::geom_line(alpha = 0.55) +
    ggplot2::geom_hline(ggplot2::aes(yintercept = truth), linetype = "dashed") +
    ggplot2::facet_wrap(~ parameter, scales = "free_y", nrow = 1) +
    ggplot2::scale_x_continuous(breaks = sort(unique(beta_plot$rep_id))) +
    ggplot2::labs(x = "Replicate", y = "Posterior mean", shape = "Status") +
    ggplot2::theme_bw(base_size = 11)
}
```

# Replicate-level diagnostics

```{r replicate-reg-table}
rep_reg_tbl <- data.frame(
  Rep = as.integer(rep_order$rep_id),
  Status = rep_order$status_label,
  `Mean count` = fmt_num(rep_order$mean_count),
  `Zero prop.` = fmt_num(rep_order$zero_prop),
  `Count CV` = fmt_num(rep_order$count_cv),
  b0 = fmt_num(rep_order$beta0_mean),
  b1 = fmt_num(rep_order$beta1_mean),
  b2 = fmt_num(rep_order$beta2_mean),
  r = fmt_num(rep_order$r_mean),
  check.names = FALSE
)
knitr::kable(rep_reg_tbl, caption = "Replicate-level count, regression, and dispersion summaries.", row.names = FALSE)
```

```{r replicate-latent-table}
rep_latent_tbl <- data.frame(
  Rep = as.integer(rep_order$rep_id),
  `log lambda RMSE` = fmt_num(rep_order$log_lambda_rmse),
  `cor(log lambda)` = fmt_num(rep_order$cor_log_lambda),
  `kappa truth CV` = fmt_num(rep_order$kappa_truth_cv),
  `kappa post CV` = fmt_num(rep_order$kappa_post_mean_cv),
  `log kappa RMSE` = fmt_num(rep_order$log_kappa_rmse),
  `cor(log kappa)` = fmt_num(rep_order$cor_log_kappa),
  check.names = FALSE
)
knitr::kable(rep_latent_tbl, caption = "Replicate-level latent-risk and kappa recovery summaries.", row.names = FALSE)
```

Replicate 1 has the largest log-lambda RMSE, but it has no numerical guard and accurate regression recovery. Replicate 8 has the lowest $\beta_1$ posterior mean among the ten replicates, but it still has stable latent-risk and kappa recovery and no guard activity. Neither replicate should be interpreted as a numerical failure.

# Dispersion-parameter recovery

Because S4C is a small-$r$ stress test, recovery of $r$ is a central metric.

```{r r-table}
r_tbl <- data.frame(
  Group = "All stable reps",
  N = as.integer(r_main$n_reps),
  `r truth` = fmt_num(r_main$r_truth_avg),
  `r mean` = fmt_num(r_main$r_mean_avg),
  `r mean SD` = fmt_num(r_main$r_mean_sd),
  `mean bias` = fmt_num(r_main$r_bias_avg),
  `mean abs. bias` = fmt_num(r_main$r_abs_bias_avg),
  `region coverage` = fmt_num(r_main$r_region_coverage_95_avg),
  check.names = FALSE
)
knitr::kable(r_tbl, caption = "Dispersion-parameter recovery for S4C continuous-time x2.", row.names = FALSE)
```

The average posterior mean of $r$ is `r fmt_num(r_main$r_mean_avg)`, compared with truth $r=3$. The model therefore learns the stronger overdispersion level, although the posterior mean is mildly above the truth on average.

```{r r-plot, fig.cap="Posterior mean of r by replicate. The dashed line shows the truth r = 3."}
if (have_ggplot2) {
  ggplot2::ggplot(rep_order, ggplot2::aes(x = rep_id, y = r_mean)) +
    ggplot2::geom_point(size = 2.1) +
    ggplot2::geom_line(alpha = 0.55) +
    ggplot2::geom_hline(yintercept = 3, linetype = "dashed") +
    ggplot2::scale_x_continuous(breaks = sort(unique(rep_order$rep_id))) +
    ggplot2::labs(x = "Replicate", y = "Posterior mean of r") +
    ggplot2::theme_bw(base_size = 11)
}
```

# Kappa recovery

S4C also requires checking the gamma-frailty component $\kappa$, because small $r$ increases observation-level overdispersion.

```{r kappa-table}
kappa_tbl <- data.frame(
  Group = "All stable reps",
  N = as.integer(kappa_main$n_reps),
  `truth CV` = fmt_num(kappa_main$kappa_truth_cv_avg),
  `reference CV` = fmt_num(kappa_main$reference_kappa_cv_avg),
  `post mean CV` = fmt_num(kappa_main$kappa_post_mean_cv_avg),
  `CV bias` = fmt_num(kappa_main$kappa_cv_bias_avg),
  `kappa RMSE` = fmt_num(kappa_main$kappa_rmse_avg),
  `log kappa RMSE` = fmt_num(kappa_main$log_kappa_rmse_avg),
  `cor(log kappa)` = fmt_num(kappa_main$cor_log_kappa_avg),
  check.names = FALSE
)
knitr::kable(kappa_tbl, caption = "Kappa recovery for S4C continuous-time x2.", row.names = FALSE)
```

The posterior-mean kappa CV is `r fmt_num(kappa_main$kappa_post_mean_cv_avg)`, close to the truth CV `r fmt_num(kappa_main$kappa_truth_cv_avg)` and much larger than the reference CV `r fmt_num(kappa_main$reference_kappa_cv_avg)`. The average correlation for log kappa is `r fmt_num(kappa_main$cor_log_kappa_avg)`. Thus, kappa-level variability is recovered well.

```{r kappa-plot, fig.cap="Truth kappa CV and posterior-mean kappa CV by replicate."}
if (have_ggplot2) {
  kcv <- rbind(
    data.frame(rep_id = rep_order$rep_id, value = rep_order$kappa_truth_cv, quantity = "Truth kappa CV"),
    data.frame(rep_id = rep_order$rep_id, value = rep_order$kappa_post_mean_cv, quantity = "Posterior kappa CV")
  )
  ggplot2::ggplot(kcv, ggplot2::aes(x = rep_id, y = value, linetype = quantity)) +
    ggplot2::geom_point(size = 2.1) +
    ggplot2::geom_line(alpha = 0.55) +
    ggplot2::scale_x_continuous(breaks = sort(unique(kcv$rep_id))) +
    ggplot2::labs(x = "Replicate", y = "Kappa CV", linetype = "Quantity") +
    ggplot2::theme_bw(base_size = 11)
}
```

# Latent temporal-risk and spatial recovery

```{r lambda-phi-table}
latent_tbl <- data.frame(
  Group = "All stable reps",
  N = as.integer(main_row$n_reps),
  `lambda RMSE` = fmt_num(main_row$lambda_rmse_avg),
  `log lambda RMSE` = fmt_num(main_row$log_lambda_rmse_avg),
  `median log RMSE` = fmt_num(main_row$log_lambda_rmse_median),
  `cor(log lambda)` = fmt_num(main_row$cor_log_lambda_avg),
  `phi RMSE` = fmt_num(main_row$phi_rmse_avg),
  `phi cor` = fmt_num(main_row$phi_cor_avg),
  check.names = FALSE
)
knitr::kable(latent_tbl, caption = "Latent temporal-risk and spatial-effect recovery for S4C continuous-time x2.", row.names = FALSE)
```

The mean log-lambda RMSE is `r fmt_num(main_row$log_lambda_rmse_avg)`, and the mean correlation for log lambda is `r fmt_num(main_row$cor_log_lambda_avg)`. This is much stronger than the latent-risk recovery observed in sparse or low-exposure stress regimes and is consistent with S4C retaining abundant count information.

```{r loglambda-plot, fig.cap="Global log-lambda RMSE by replicate."}
if (have_ggplot2) {
  ggplot2::ggplot(rep_order, ggplot2::aes(x = rep_id, y = log_lambda_rmse)) +
    ggplot2::geom_point(size = 2.1) +
    ggplot2::geom_line(alpha = 0.55) +
    ggplot2::scale_x_continuous(breaks = sort(unique(rep_order$rep_id))) +
    ggplot2::labs(x = "Replicate", y = "Log-lambda RMSE") +
    ggplot2::theme_bw(base_size = 11)
}
```

# Numerical guards

```{r guard-table}
guard_main <- guard_agg[guard_agg$subset == "stable_plus_soft_warning", , drop = FALSE]
if (nrow(guard_main) == 0) guard_main <- guard_agg[guard_agg$subset == "all_sampled_lambda", , drop = FALSE]
guard_tbl <- data.frame(
  Group = "All stable reps",
  N = as.integer(guard_main$n_reps),
  `Beta guards` = fmt_int(guard_main$beta_guard_total),
  `Kappa guards` = fmt_int(guard_main$kappa_guard_total),
  `Lambda-input guards` = fmt_int(guard_main$lambda_input_guard_total),
  `Lambda-output guards` = fmt_int(guard_main$lambda_output_guard_total),
  check.names = FALSE
)
knitr::kable(guard_tbl, caption = "Numerical guard summary for S4C continuous-time x2.", row.names = FALSE)
```

There were no beta guards, no kappa guards, no lambda-input guards, and no lambda-output guards. This confirms that the continuous-time S4C result is not a numerical failure scenario.

# Comparison with the S3 control

```{r comparison-table}
if (!is.null(comparison) && nrow(comparison) > 0) {
  comp <- comparison
  comp$Scenario <- comp$scenario
  comp$Scenario[comp$scenario == "S4C overdispersion"] <- "S4C overdispersion"
  comp$Scenario[comp$scenario == "S3 control"] <- "S3 control"
  comp$Group <- comp$subset
  comp$Group[comp$subset == "stable+soft"] <- "All stable"
  comp$Group[comp$subset == "all_control_reps"] <- "All"
  comp_tbl <- data.frame(
    Scenario = comp$Scenario,
    Group = comp$Group,
    N = as.integer(comp$n_reps),
    `Mean count` = fmt_num(comp$mean_count_avg, 2),
    `Zero prop.` = fmt_num(comp$zero_prop_avg),
    `b1 mean` = fmt_num(comp$beta1_mean_avg),
    `b2 mean` = fmt_num(comp$beta2_mean_avg),
    `r mean` = fmt_num(comp$r_mean_avg),
    `log RMSE` = fmt_num(comp$log_lambda_rmse_avg),
    `cor(log)` = fmt_num(comp$cor_log_lambda_avg),
    check.names = FALSE
  )
  knitr::kable(comp_tbl, caption = "Comparison of S3 control and S4C continuous-time strong-overdispersion stress.", row.names = FALSE)
}
```

Relative to the S3 control, S4C preserves regression-slope recovery while increasing count variability and kappa variability. The S4C log-lambda RMSE is higher than the S3 control but still much smaller than the failure-mode values observed in sparse-count stress settings.

# Interpretation of the stress mechanism

The continuous-time S4C results separate three phenomena.

First, strong overdispersion increases observation-level variability without creating global sparsity. The average count remains high and the zero proportion remains very low.

Second, the model learns the smaller dispersion parameter and kappa-level variability. The posterior mean of $r$ is close to the truth, and the posterior-mean kappa CV is close to the truth kappa CV.

Third, under the continuous-time $x_2$ design, regression slopes remain recoverable. This is materially different from a coefficient-fragility or numerical-failure scenario. The dominant S4C message is therefore positive: strong overdispersion is mostly manageable when count information remains abundant.

# Practical implications

The S4C continuous-time $x_2$ results suggest several practical implications for MSSTNB applications.

1. Strong overdispersion should be diagnosed separately from sparse counts and low exposure. It increases variability, but it does not necessarily destroy information when counts remain abundant.
2. Dispersion recovery should be explicitly reported. In this setting, the posterior mean of $r$ is close to the truth $r=3$.
3. Kappa diagnostics should distinguish variability-level recovery from exact cell-level recovery. Here, kappa CV and log-kappa correlation are both informative.
4. Fit-status classification remains useful, but in this continuous-time S4C setting it mainly documents stability.
5. Real applications with strong overdispersion may still need careful posterior checks for $r$, $\kappa$, and $\lambda$, even if regression coefficients look accurate.

# Final conclusion

Scenario 4C with continuous-time $x_2$ confirms that reducing the negative-binomial dispersion parameter from $r=15$ to $r=3$ creates a strong-overdispersion stress regime without creating global sparsity. The data have high average counts, low zero proportions, increased count CV, increased variance-to-mean ratio, and substantially higher kappa variability than the reference setting.

Under this regime, all ten fixed-gamma sampled-$\lambda$ fits are stable and no numerical guards occur. The regression slopes are accurately recovered: the posterior means are `r fmt_num(b1_main$mean_post_mean)` for $\beta_1$ and `r fmt_num(b2_main$mean_post_mean)` for $\beta_2$, compared with truths 0.5 and -0.4. The model also recovers the stronger overdispersion level, with average posterior mean $r = `r fmt_num(r_main$r_mean_avg)`$, and recovers kappa-level variability, with posterior-mean kappa CV `r fmt_num(kappa_main$kappa_post_mean_cv_avg)` compared with truth CV `r fmt_num(kappa_main$kappa_truth_cv_avg)`.

Therefore, S4C should be interpreted as a strong-overdispersion stress result with successful recovery under abundant count information, not as a regression-slope failure or a beta/r/kappa implementation failure.

# Recommended reporting language

Scenario 4C evaluated the MSSTNB model under strong negative-binomial overdispersion while using the same continuous-time $x_2$ covariate design as the main dynamic simulations. The dispersion parameter was reduced from $r=15$ to $r=3$, increasing count variability and kappa-level variability without creating a sparse-count regime. In the fixed-gamma sampled-$\lambda$ fit, all ten replicates were stable and no numerical guards occurred. Both regression slopes were accurately recovered, the posterior mean of $r$ was close to the truth, and kappa-level variability was recovered well. These results indicate that strong overdispersion is largely manageable for the MSSTNB model when count information remains abundant, although latent-risk and frailty diagnostics should still be reported explicitly.
